Cosmological Evidences of Dark Matter through the CMB

### 2019-02-22 paper/main.tex
dark energy and curvature

%The recent measurements of the Cosmic Microwave Background (CMB) radiation allowed us to infer the presence of Dark Matter in the Universe. In this summary we will explain qualitatively how the presence of Dark Matter influences the CMB anisotropies.
\pagebreak
\section{From the Discovery of the CMB to the Planck mission}
%
%
%
\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{Horn_Antenna.jpeg}
\caption{The 15 meter Holmdel horn antenna at Bell Telephone Laboratories in Holmdel, New Jersey.
(By NASA - Great Images in NASA Description, Public Domain, \url{https://commons.wikimedia.org/w/index.php?curid=6463768})}
\label{antenna}
\end{center}
\end{figure}
%
%
%
In the 1960s Penzias and Wilson measured an isotropic and constant with time strong signal coming from the sky by using a horn-reflector radio antenna (Figure(\ref{antenna})). 
They tried everything they could do to reduce the “noise” in their system, but the signal remained.\\
In 1965 Arno Penzias and Robert Wilson published the paper: \emph{A measurement of excess antenna temperature at
4080 Mc/s}, where they reported the measurements of an isotropic, unpolarized, and free from seasonal variations excess noise temperature
\citep{penziasMeasurementExcessAntenna1965}.
The explanation of that signal was given by Robert Dicke and his research group at Princeton University: such signal as the relic of an early, hot, dense, and opaque state of the universe\footnote{The existence of the cosmic background radiation had actually been predicted by the physicist George Gamow and his colleagues in 1948} \citep{bucherPhysicsCosmicMicrowave2015}
. \\
The discovery of the CMB created a big excitement and a new era of cosmology began.
Measuring the CMB spectrum and its deviation from the black body spectrum was the new challange.
\\
First of all, the photons of the CMB can be absorbed by the Earth's atmosphere, because the energy per CMB photon (approximately $\sim 6 \times 10 ^{-4}$ eV) is comparable to the energy of vibration or rotation for a small molecule (of water for instance). 
In fact, microwaves with wavelengths shorter than $\lambda \sim 3$ cm are strongly absorbed by water molecules \citep{RydenIntroCosmoPdf}.
This problem could be overcomed by observing the CMB in different bands and from locations where the atmospheric humidity is low, at high altitudes and at low temperatures.
However, the best way to meseaure the CMB spectrum is to use satellites.\\
The first sattellite that was launched to observe the CMB was COBE. 
It did provide a convincing first detection of the CMB anisotropy, and it played a crucial role in determining the viability of the different cosmological models at that time.
\\
After that, WMAP and PLanck space missions  increased the accuracy of the measurements Figure (\ref{cobe_wmap_planck}).\\
The COBE, WMAP and Planck missions measured the CMB anisotropies with increasing accuracy. COBE, the first CMB satellite, measured fluctuations to scales of 7$^\circ$. WMAP was able to measure resolutions down to 0.3$^\circ$ in five different frequency bands, with Planck measuring all the way down to just 5 arcminutes (0.07$^\circ$) in nine different frequency bands in total.\\
Nevertheless the satellites improved their accuracy, many other sources of noise were encountered such as the light coming from the center of our galaxy, other stars and other objects in the universe, in addition to the relative motion of the satellite with respect to the CMB (Figure (\ref{cobe_map})).\\ 
Finally, observations and statistical analysis showed that the CMB spectrum is close to a black body spectrum up to fluctuations of order of $10^{-5}$ K (Figure (\ref{cobe_blackbody})).\\
% maybe last sentence in the introduction with the image

@@

\section{Content of the Universe} 

@@

\section{Comments on the results and plots of the CMB spectrum}
The four density parameters $\Omega$ encode the energy-matter content of the Universe and they determine the evolution of space-time through the Friedmann equations. 
In section 4 we have outlined how we expect matter to behave at large and small scales.
We want now to link the information contained in the fluctuations of the CMB photons to the density parameters.\\
The behavior of the primordial plasma is imprinted in the CMB photons, in fact, we observe the density fluctuations as anisotropies in the sky map tempearature.
In order to relate the fluctuations at different scales and the deviations from the average temperature $2.7$ K on the CMB map, it is necessary to decompose the temperature deviation $T(\theta, \phi)$ into a set of orthonormal functions on the sky.
The spherical harmonics $Y^{m} _{l} (\theta, \phi)$ are suitable for this.\\
So if $T(\theta,\phi)$ is the temperature at position $(\theta,\phi)$ on the sky map,
we can write:
\begin{equation}
T(\theta, \phi) = \sum _{l, m} a_{lm} Y^{m} _l (\theta, \phi)
\end{equation}
Due to the orthonormality of the spherical harmonics, we obtain the coeaficients:
\begin{equation}
a_{lm} = \int _0 ^{2 \pi} \dd \phi \int _0  ^{\pi} \dd \theta \,
\sin( \theta) \, T(\theta, \phi) \, Y^{m} _l (\theta, \phi)
 \end{equation}
We define the power spectrum of the temperature map as 
 \begin{equation}
 C_l = \dfrac{1}{2l+1} \sum _{m=-l} ^{l} |a_{lm}|^2
 \end{equation}
Since $l \sim 1/ \theta$, for small $l$ we are considering big scales of the sky map, viceversa for big $l$.
As we consider increasing $l$ we are looking at fluctuations in the primordial plasma that have a decreasing  scale and thus wavelength.\\
The measured temperature power spectrum is shown in Figure(\ref{temp_pow_spect})
%
%
%
\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{planck2018}
\caption{Planck 2018 temperature power spectrum. $\mathcal{D}^{TT}_l = l(l+1)C_l$ (arXiv:1807.06209v1)}
\label{temp_pow_spect}
\end{figure}
%
%
%
For large scales, small $l<30$, the power spectrum is constant and oscillations do not take place.
Therefore, the Sachs-Wolfe effect is dominant on large scales where the oscillations have too large wavelength to complete an entire oscillations before recombination and they experience mainly the gravitational redshift.\\
For large $l$ the acoustic oscillations are dominant.
The first peak is at $l \approx 200$ approximately $1 ^o$ on the sky, it represents how the tiny variations in the fluid density gave rise to a sound wave which collapsed exactly at the time of recombination.
The position of the peak tells us at which scale the first total compression of the primordial plasma happened.
The second peak is linked to those oscillations that compressed and expanded, and they reached the maximal rarefaction of the plasma at recombinatio.
The third peak is related to the oscillations that collapsed expanded and recollapsed, and so on for the higher peaks.\\
Notice that as we consider higher and higher $l$, i.e. smaller scales, the damping effects become relevant, especially for  $l>10^{3}$.

\subsection{Dark Energy and Curvature}

Position of peaks mainly sensitive to curvature
Shapes fixed by the physical density of matter and baryons
Missing or "dark energy" plays a small roll in the position of peaks
As advertised, the position of that first peak in the power spectrum of the anisotropies, and indeed all of the peaks, depend sensitively on the spatial curvature of the universe.   As the curvature of the universe decreases (and in fact becomes negative in the yellow curve below)\\
the peaks move to smaller angles (higher multipole l) while preserving their shape.  Cosmologists fraction of the critical density in matter Omegam so that as 1-Omegamincreases from zero, the universe becomes increasingly negatively curved if there is no other forms of "missing energy" that we've missed in our accounting.

Caveats:

In the blue curve, we assume that there is indeed a form of "missing" or dark energy in a cosmological constant that makes the universe flat despite a sub-critical density of matter.  We see that the positions mainly care about the curvature of the universe but there is a small shift to larger angular scales (lower multipoles) through the introduction of a cosmological constant.  This is because the cosmological constant produces a small change in the distance light can travel since recombination, a fact that is related to its well known effect on the age of the universe.

Note that in the above figure we assume that the physical density of the matter (Omegamh2) is fixed.  This is different from saying its fraction of the critical density (Omegam) is fixed since the Hubble constant (h), or expansion rate today, enters into the definition of the critical density.  Given our current uncertainties about the physical density of the matter, the distance that sound can travel by recombination is currently uncertain, a fact related to the matter's effect on the age of the universe at recombination.  Likewise we have assumed that the physical density of the baryons  (Omegabh2) is fixed.  Baryons lower the speed of sound in the medium and hence also affect the distance sound can travel.  Luckily both of these quantities dramatically change the shape of the peaks as we shall see.  They will be measured once the higher peaks are detected and will no longer confuse the measurement of the curvature.   As it turns out, the sound horizon is not a standard ruler but rather a standardizeable ruler.
\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{curvature_de}
\caption{}

\end{center}
\label{DE_curv}
\end{figure}

>>>>

%The recent measurements of the Cosmic Microwave Background (CMB) radiation allowed us to infer the presence of Dark Matter in the Universe. In this summary we will explain qualitatively how the presence of Dark Matter influences the CMB anisotropies.
\pagebreak
\section{From the Discovery of the CMB to the Planck Mission}
%
%
%
\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{Horn_Antenna.jpeg}
\caption{The 15 meter Holmdel horn antenna at Bell Telephone Laboratories in Holmdel, New Jersey.
(By NASA - Great Images in NASA Description, Public Domain, \url{https://commons.wikimedia.org/w/index.php?curid=6463768})}
\label{antenna}
\end{center}
\end{figure}
%
%
%
In the 1960s Penzias and Wilson measured an isotropic and constant with time strong signal coming from the sky by using a horn-reflector radio antenna (Figure(\ref{antenna})). 
They tried everything they could do to reduce the “noise” in their system, but the signal remained.\\
In 1965 Arno Penzias and Robert Wilson published the paper: \emph{A measurement of excess antenna temperature at
4080 Mc/s}, where they reported the measurements of an isotropic, unpolarized, and free from seasonal variations excess noise temperature
\citep{penziasMeasurementExcessAntenna1965}.
The explanation of that signal was given by Robert Dicke and his research group at Princeton University: such signal as the relic of an early, hot, dense, and opaque state of the universe\footnote{The existence of the cosmic background radiation had actually been predicted by the physicist George Gamow and his colleagues in 1948} \citep{bucherPhysicsCosmicMicrowave2015}
. \\
The discovery of the CMB created a big excitement and a new era of cosmology began.
Measuring the CMB spectrum and its deviation from the black body spectrum was the new challange.
\\
First of all, the photons of the CMB can be absorbed by the Earth's atmosphere, because the energy per CMB photon (approximately $\sim 6 \times 10 ^{-4}$ eV) is comparable to the energy of vibration or rotation for a small molecule (of water for instance). 
In fact, microwaves with wavelengths shorter than $\lambda \sim 3$ cm are strongly absorbed by water molecules \citep{RydenIntroCosmoPdf}.
This problem could be overcomed by observing the CMB in different bands and from locations where the atmospheric humidity is low, at high altitudes and at low temperatures.
However, the best way to meseaure the CMB spectrum is to use satellites.\\
The first sattellite that was launched to observe the CMB was COBE. 
It did provide a convincing first detection of the CMB anisotropy, and it played a crucial role in determining the viability of the different cosmological models at that time.
\\
After that, WMAP and PLanck space missions  increased the accuracy of the measurements Figure (\ref{cobe_wmap_planck}).\\
The COBE, WMAP and Planck missions measured the CMB anisotropies with increasing accuracy. COBE, the first CMB satellite, measured fluctuations to scales of 7$^\circ$. WMAP was able to measure resolutions down to 0.3$^\circ$ in five different frequency bands, with Planck measuring all the way down to just 5 arcminutes (0.07$^\circ$) in nine different frequency bands in total.\\
Nevertheless the satellites improved their accuracy, many other sources of noise were encountered such as the light coming from the center of our galaxy, other stars and other objects in the universe, in addition to the relative motion of the satellite with respect to the CMB (Figure (\ref{cobe_map})).\\ 
Finally, observations and statistical analysis showed that the CMB spectrum is close to a black body spectrum up to fluctuations of order of $10^{-5}$ K (Figure (\ref{cobe_blackbody})).\\
% maybe last sentence in the introduction with the image

@@

\section{Energy-Matter Content of the Universe} 

@@

\section{Comments on the results and plots of the CMB spectrum}
We have outlined how we expect matter to behave at large and small scales.
Let us now see if our predictions are compatible with the measured temperature power spectrum.
The behavior of the primordial plasma is imprinted in the CMB photons, in fact, we observe the density fluctuations as anisotropies in the sky map tempearature.
In order to relate the fluctuations at different scales and the deviations from the average temperature $2.7$ K on the CMB map, it is necessary to decompose the observed temperature deviation $T(\theta, \phi)$ into a set of orthonormal functions on the sky.
The spherical harmonics $Y^{m} _{l} (\theta, \phi)$ are suitable for this.\\
So if $T(\theta,\phi)$ is the temperature at position $(\theta,\phi)$ on the sky map,
we can write:
\begin{equation}
T(\theta, \phi) = \sum _{l, m} a_{lm} Y^{m} _l (\theta, \phi)
\end{equation}
Due to the orthonormality of the spherical harmonics, we obtain the coeaficients:
\begin{equation}
a_{lm} = \int _0 ^{2 \pi} \dd \phi \int _0  ^{\pi} \dd \theta \,
\sin( \theta) \, T(\theta, \phi) \, Y^{m} _l (\theta, \phi)
 \end{equation}
We define the power spectrum of the temperature map as 
 \begin{equation}
 C_l = \dfrac{1}{2l+1} \sum _{m=-l} ^{l} |a_{lm}|^2
 \end{equation}
Since $l \sim 1/ \theta$, for small $l$ we are considering big scales of the sky map, viceversa for big $l$.
As we consider increasing $l$ we are looking at fluctuations in the primordial plasma that have a decreasing  scale and thus a smaller wavelength.\\
The measured temperature power spectrum is shown in Figure(\ref{temp_pow_spect}).
%
%
%
\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{planck2018}
\caption{Planck 2018 temperature power spectrum. $\mathcal{D}^{TT}_l = l(l+1)C_l$ (arXiv:1807.06209v1)}
\label{temp_pow_spect}
\end{figure}
%
%
%
For large scales, small $l<30$, the power spectrum is constant and oscillations do not take place.
Therefore, the Sachs-Wolfe effect is dominant on large scales where the oscillations have too large wavelength to complete an entire oscillations before recombination and they experience mainly the gravitational redshift.\\
For large $l$ the acoustic oscillations are dominant.
The first peak is at $l \approx 200$ approximately $1 ^o$ on the sky, it represents how the tiny variations in the fluid density gave rise to a sound wave which collapsed exactly at the time of recombination.
The position of the peak tells us at which scale the first total compression of the primordial plasma happened.
The second peak is linked to those oscillations that compressed and expanded, and they reached the maximal rarefaction of the plasma at recombinatio.
The third peak is related to the oscillations that collapsed expanded and recollapsed, and so on for the higher peaks.\\
Notice that as we consider higher and higher $l$, i.e. smaller scales, the damping effects become relevant, especially for  $l>10^{3}$.\\

The four density parameters $\Omega$ encode the energy-matter content of the Universe and they determine the evolution of space-time through the Friedmann equations. 
We want now to link the information contained in the measured fluctuations of the CMB to the density parameters through the observed power spectrum.\\

\subsection{Dark Energy and Curvature}
The cosmological constant $\Lambda$ was introduced in the Einstein's field equations as a possible parameter of our cosmological model.
It turns out from the measurements of the Hubble constant $H_0$ and distant supernovae that the Universe is not only expanding but accelerating.
Therefore, from the Friedmann equations we know that the cosmological constant plays a crucial role in the expansion and in the future evolution of the Universe.
This mysterious energy that leads to the expansion of the Univers is called Dark Energy.\\
Form Figure(\ref{DE_curv} b) we can notice that the dark energy density $\Omega_{\Lambda0}$ plays a small role in the position of the peaks. 
In fact, the  cosmological constant produces a small change in the distance light can travel since recombination and this influences the scale at which we measure the peaks.\\
The position of the peaks is instead particularly sensitive to the cruvature of the Universe, as show in Figure(\ref{DE_curv}a).
The curvature determines how anlges are deformed.
By measuring the position of the first peak we know the angle $\theta \sim 1/l$ of the oscillations.
If I multiply by the angle by the distance (specifically the angular distance) I obtain the size of the oscillations for Euclidean geometry $k =0$.
The first peak represents the first complete compression of the photon-bryon fluid, so by multplying the sound speed of the fluid times the time of the collapse times the expansion factor I obtain the size of the oscillations and I can compare it with the estimated size in order to check the curvature parameter.
	TO BE EXPLAINED DECENTLY

% Given our current uncertainties about the physical density of the matter, the distance that sound can travel by recombination is currently uncertain, a fact related to the matter's effect on the age of the universe at recombination.  Likewise we have assumed that the physical density of the baryons  (Omegabh2) is fixed.  Baryons lower the speed of sound in the medium and hence also affect the distance sound can travel.  Luckily both of these quantities dramatically change the shape of the peaks as we shall see.  They will be measured once the higher peaks are detected and will no longer confuse the measurement of the curvature.   As it turns out, the sound horizon is not a standard ruler but rather a standardizeable ruler.
\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{curvature_de}
\end{center}
\caption{}
\label{DE_curv}
\end{figure}
